---
title: 'Reproducible Research: Project 01'
author: "Thiago Veras"
date: "July 13, 2015"
output: html_document
---

### Dataset: Activity monitoring data
#### The variables included in this dataset are:
- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)   
- date: The date on which the measurement was taken in YYYY-MM-DD format   
- interval: Identifier for the 5-minute interval in which measurement was taken   

#### The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

### Loading and preprocessing the data
#### 1. Loading data
```{r loading_data}
# Loading the data
raw_data <- read.csv("activity.csv")

# Printing the head of the raw data
head(raw_data)
```

#### 2. Preprocessing the data
```{r preprocessing_data}
# Transforming the date attribute to an actual date format
raw_data$date <- as.POSIXct(raw_data$date, format="%Y-%m-%d")

# Computing the weekdays from the date attribute
raw_data <- data.frame(date=raw_data$date, 
                           weekday=tolower(weekdays(raw_data$date)), 
                           steps=raw_data$steps, 
                           interval=raw_data$interval)

# Create the activity data.frame
activity_data <- data.frame(date=raw_data$date, 
                            weekday=raw_data$weekday, 
                            interval=raw_data$interval,
                            steps=raw_data$steps)

# Printing the head of activity data.frame
head(activity_data)
```

### What is mean total number of steps taken per day?
#### 1. Calculating the total number of steps taken per day
```{r}
# Aggregating steps by date
aggregated_activity_na_rm <- aggregate(activity_data$steps, 
                                        by=list(activity_data$date), 
                                        FUN=sum, na.rm=TRUE)
# Rename the attributes
names(aggregated_activity_na_rm) <- c("date", "total_steps")
```

#### 2. Making a histogram of the total number of steps taken each day
```{r}
# Importing the ggplot2 library
library(ggplot2)

# Ploting the graph
ggplot(aggregated_activity_na_rm, aes(x=total_steps)) + 
    geom_histogram(binwidth=1000, breaks=seq(from=0, to=25000, by=2500), colour="black", fill="darkred") +
    ylim(0,20) +
    ggtitle("Histogram of the total of steps taken each day\n(NA removed)") +
    xlab("Total of Steps") +
    ylab("Frequency")
``` 

#### 3. Calculating and reporting the mean and median of the total number of steps taken per day
```{r}
# Calculating the mean
mean(aggregated_activity_na_rm$total_steps, na.rm=TRUE)

# Calculating the median
median(aggregated_activity_na_rm$total_steps, na.rm=TRUE)
```

### What is the average daily activity pattern?
#### 1. Making a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
# Importing the ggplot2 library
library(ggplot2)

# Aggregating average of steps by interval
average_activity_by_interval <- aggregate(x=list(steps=activity_data$steps), 
                                   by=list(interval=activity_data$interval),
                                   FUN=mean, na.rm=TRUE)

ggplot(data=average_activity_by_interval, aes(x=interval, y=steps)) +
    geom_line(binwidth=100, breaks=seq(from=0, to=3000, by=500), colour="darkred") +
    ylim(0,250) +
    xlab("5-minute interval") +
    ylab("Average across all days")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
average_activity_by_interval[which.max(average_activity_by_interval$steps),]
```

### Imputing missing values
#### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
#### 1. Calculating and reporting the total number of missing values in the dataset.
```{r}
total_na <- sum(is.na(activity_data$steps))
total_na
```

#### 2. Devising a strategy for filling in all of the missing values in the dataset. 
The strategy = The missing values will be filled in with mean value for that 5-minute interval.

#### 3. Creating a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# Finding the NA positions
na_positions <- which(is.na(activity_data$steps))

# Creating a vector of means
mean_vector <- rep(mean(activity_data$steps, na.rm=TRUE), times=length(na_positions))

# Replacing the NAs by the means
activity_data[na_positions, "steps"] <- mean_vector

# Printing the head of activity data
head(activity_data)
```

#### 4. Making a histogram of the total number of steps taken each day.

```{r}
# Importing the ggplot2 library
library(ggplot2)

# Aggregating steps by date
aggregated_activity_by_day <- aggregate(activity_data$steps, 
                                        by=list(activity_data$date), 
                                        FUN=sum, na.rm=TRUE)

# Rename the attributes
names(aggregated_activity_by_day) <- c("date", "total_steps")

# Ploting the graph
ggplot(aggregated_activity_by_day, aes(x=total_steps)) + 
    geom_histogram(binwidth=1000, breaks=seq(from=0, to=25000, by=2500), colour="black", fill="darkred") +
    ylim(0,30) +
    ggtitle("Histogram of the total of steps taken each day\n(NA replaced by mean value)") +
    xlab("Total of Steps") +
    ylab("Frequency")
```

#### Calculating and reporting the mean and median total number of steps taken per day. 
```{r}   
# Calculating the mean
mean(aggregated_activity_by_day$total_steps, na.rm=TRUE)

# Calculating the median
median(aggregated_activity_by_day$total_steps, na.rm=TRUE)
```

#### Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The mean and median values are higher after filled in missing data. 
The reason is that in the original data, there are some days with `steps` values `NA` for any `interval`. 
The total number of steps taken in such days are set to 0s by default. 
However, after replacing missing `steps` values with the mean `steps` of associated `interval` value, these 0 values are removed from the histogram of total number of steps taken each day.

### Are there differences in activity patterns between weekdays and weekends?
#### 1. Creating a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
head(activity_data)

# Computing the day type (weekend or weekday)
activity_data <- cbind(activity_data, 
                  daytype=ifelse(activity_data$weekday == "saturday" | 
                                 activity_data$weekday == "sunday", 
                                 "weekend", "weekday"))

head(activity_data)

```

#### 2. Making a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r}
averages_daytype <- aggregate(steps ~ interval + daytype, data=activity_data, mean)
                    ggplot(averages_daytype, aes(interval, steps)) + 
                    geom_line(binwidth=100, breaks=seq(from=0, to=3000, by=500), colour="darkred") +
                    ylim(0,250) + 
                    facet_grid(daytype ~ .) +
                    xlab("5-minute interval") + 
                    ylab("averaged across all days selected")
                    
```
