---
title: 'Reproducible Research: Project 01'
author: "Thiago Veras"
date: "July 13, 2015"
output: html_document
---

### Dataset: Activity monitoring data
#### The variables included in this dataset are:
- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)   
- date: The date on which the measurement was taken in YYYY-MM-DD format   
- interval: Identifier for the 5-minute interval in which measurement was taken   

#### The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

### Loading and preprocessing the data
#### Loading data
```{r loading_data}
# Loading the data
raw_data <- read.csv("activity.csv")

# Printing the head of the raw data
head(raw_data)
```

#### Preprocessing the data
```{r preprocessing_data}
# Transforming the date attribute to an actual date format
raw_data$date <- as.POSIXct(raw_data$date, format="%Y-%m-%d")

# Computing the weekdays from the date attribute
raw_data <- data.frame(date=raw_data$date, 
                           weekday=tolower(weekdays(raw_data$date)), 
                           steps=raw_data$steps, 
                           interval=raw_data$interval)

# Create the activity data.frame
activity_data <- data.frame(date=raw_data$date, 
                            weekday=raw_data$weekday, 
                            interval=raw_data$interval,
                            steps=raw_data$steps)

# Printing the head of activity data.frame
head(activity_data)
```

### What is mean total number of steps taken per day?
```{r}
# Importing the ggplot2 library
library(ggplot2)

# Aggregating steps by date
aggregated_activity_na_rm <- aggregate(activity_data$steps, 
                                        by=list(activity_data$date), 
                                        FUN=sum, na.rm=TRUE)

# Rename the attributes
names(aggregated_activity_na_rm) <- c("date", "total_steps")

# Ploting the graph
ggplot(aggregated_activity_na_rm, aes(x=total_steps)) + 
    geom_histogram(binwidth=1000, breaks=seq(from=0, to=25000, by=2500), colour="black", fill="darkred") +
    ylim(0,20) +
    ggtitle("Histogram of the total of steps taken each day\n(NA removed)") +
    xlab("Total of Steps") +
    ylab("Frequency")
    
# Calculating the mean
mean(aggregated_activity_na_rm$total_steps, na.rm=TRUE)

# Calculating the median
median(aggregated_activity_na_rm$total_steps, na.rm=TRUE)
```

### What is the average daily activity pattern?
```{r}
# Importing the ggplot2 library
library(ggplot2)

# Aggregating average of steps by interval
average_activity_by_interval <- aggregate(x=list(steps=activity_data$steps), 
                                   by=list(interval=activity_data$interval),
                                   FUN=mean, na.rm=TRUE)

ggplot(data=average_activity_by_interval, aes(x=interval, y=steps)) +
    geom_line(binwidth=100, breaks=seq(from=0, to=3000, by=500), colour="darkred") +
    ylim(0,250) +
    xlab("5-minute interval") +
    ylab("Average of steps")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
average_activity_by_interval[which.max(average_activity_by_interval$steps),]
```

### Imputing missing values
#### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
total_na <- sum(is.na(activity_data$steps))
total_na
```

All of the missing values are filled in with mean value for that 5-minute interval.
```{r}
# Finding the NA positions
na_positions <- which(is.na(activity_data$steps))

# Creating a vector of means
mean_vector <- rep(mean(activity_data$steps, na.rm=TRUE), times=length(na_positions))

# Replacing the NAs by the means
activity_data[na_positions, "steps"] <- mean_vector

# Printing 
head(activity_data)
```

Now, using the filled data set, let's make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
# Importing the ggplot2 library
library(ggplot2)

# Aggregating steps by date
aggregated_activity_by_day <- aggregate(activity_data$steps, 
                                        by=list(activity_data$date), 
                                        FUN=sum, na.rm=TRUE)

# Rename the attributes
names(aggregated_activity_by_day) <- c("date", "total_steps")

# Ploting the graph
ggplot(aggregated_activity_by_day, aes(x=total_steps)) + 
    geom_histogram(binwidth=1000, breaks=seq(from=0, to=25000, by=2500), colour="black", fill="darkred") +
    ylim(0,30) +
    ggtitle("Histogram of the total of steps taken each day\n(NA replaced by mean value)") +
    xlab("Total of Steps") +
    ylab("Frequency")
    
# Calculating the mean
mean(aggregated_activity_by_day$total_steps, na.rm=TRUE)

# Calculating the median
median(aggregated_activity_by_day$total_steps, na.rm=TRUE)
```

Mean and median values are higher after imputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

### Are there differences in activity patterns between weekdays and weekends?
```{r}
head(activity_data)

# Computing the day type (weekend or weekday)
activity_data <- cbind(activity_data, 
                  daytype=ifelse(activity_data$weekday == "saturday" | 
                                 activity_data$weekday == "sunday", 
                                 "weekend", "weekday"))

head(activity_data)

```

Now, let's make a panel plot containing plots of average number of steps taken
on weekdays and weekends.
```{r}
averages_weekday <- aggregate(steps ~ interval + weekday, data=activity_data, mean)
                    ggplot(averages_weekday, aes(interval, steps)) + 
                    geom_line(binwidth=100, breaks=seq(from=0, to=3000, by=500), colour="darkred") +
                    ylim(0,400) + 
                    facet_grid(weekday ~ .) +
                    xlab("5-minute interval") + 
                    ylab("Total of steps")


averages_daytype <- aggregate(steps ~ interval + daytype, data=activity_data, mean)
                    ggplot(averages_daytype, aes(interval, steps)) + 
                    geom_line(binwidth=100, breaks=seq(from=0, to=3000, by=500), colour="darkred") +
                    ylim(0,250) + 
                    facet_grid(daytype ~ .) +
                    xlab("5-minute interval") + 
                    ylab("Total of steps")
                    
```
